Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-07-08 evaluation.tex
Edits.

To explore the space of correct isolation levels for different transactions, we  also  ran a number  of experiments using  a real transaction OLTP workload   based on TPC-C benchmark.  
 The  workload  involves  a mix of five TPC-C transaction types: 
\C{New\_Order},  \C{Payment},  \C{Delivery}, \C{Stock Level} and  \C{Order Status}, 
operating on top of  a $10$-warehouse database.  
Each warehouse has  $10$ districts, each serving $3000$ customers. 
TPC-C  specification model  requires  maintaining multiple   consistency conditions.  We focus  on   four conditions, named $I_1$, $I_2$, $I_3$, and $I_4$.   
Consistency condition $I_1$  requires that  sales of each  warehouse  equals to the  sum of    its  district's sales.  Conditions $I_2$ and $I_3$  enforce  
 sequentially unique IDs assigned to  orders placed in the same   warehouse and district. Condition $I_4$ requires 
 the  number of  item orders   in each district 
  reflexes  the district's roll-up. 
Given the consistency conditions,  Table  \ref{tab:tpcc-eval} gives the isolation level  for the \C{New\_Order},  \C{Payment},  \C{Delivery} transactions using the  MySQL and PostgreSQL databases   to maintain   each consistency condition. Note,   \C{Stock Level} and  \C{Order Status}   are read-only transactions can be executed under RC.   
Recall that the \C{New\_Order} transaction places a new order for
 items from a particular district in a warehouse.  
The \C{Payment} transaction  processes  a payment for a particular customer and updates the customer, district, and warehouse tables. 
 The  \C{Delivery}  transaction processes  orders for a particular   district at  a warehouse. 
 Our experimental results  show that  the concurrent executions of   \C{New\_Order}   
    and  \C{Delivery} transactions under a weak isolation level (i.e.,  RC) is anomalous, i.e., it may violate the consistency conditions $I_2$ and $I_3$. Depending on the isolation guarantees provided  by  underlying database,   the stronger  isolation level is required. For instance, the isolation level RR in PostgreSQL  is sufficient to preserve the consistency  conditions while  operating on MySQL database  requires  these transaction to be serialized.

 	\begin{table*}[t]\small
	\centering
\begin{tabular}{|c|c|c|c|}
		\hline
	\textbf{Transaction}   & \textbf{Invariant} 
	& \textbf{MySQL-Isolation} & \textbf{PostgreSQL-Isolation} \\ 
	\hline
	\multirow{4}{*}{\C{New\_Order} }  & $I_1$ 
	& RC &  RC\\ 
	&  $I_2$ &SER & RR \\
	&  $I_3$ & SER  &  RR  \\
	& $I_4$ & RC & RC   \\
	\hline
	\multirow{4}{*}{\C{Payment}}  & $I_1 $ 
	& RC &  RC\\ 
	&  $I_2$  &RC & RC \\
	&  $I_3 $ & RC  &  RC  \\
	& $I_4$  & RC & RC   \\
	\hline
	\multirow{4}{*}{\C{Delivery}}  & $I_1$  
	& RC &  RC \\ 
	&  $I_2$ &SER & RR \\
	&  $I_3$ & SER  &  RR \\
	& $I_4$  & RC & RC   \\
	\hline
\end{tabular}
	\caption{ Isolation level required by TPC-C transactions. }
	\label{tab:tpcc-eval}
\end{table*}

>>>>

We evaluate \tool on two different applications: 1. Courseware: a
course registration system taken from~\cite{gotsmanpopl16}, and
% 2. RUBiS: an eBay-like auction site~\cite{rubis} that allows users to
% bid for items on sale, and pay for items from a wallet modeled after a
% bank account, and 
2. The TPC-C benchmark.

\textbf{Courseware} Courseware application allows new courses to be
added (\C{add\_course} transaction), and new students to be
registered (\C{register\_student}) into the system. A registered
student can enroll (\C{enroll}) in one of the existing courses,
provided that the enrollments haven't already exceeded the course
capacity (\C{c\_capacity}). A course with no enrollments can be
canceled (\C{cancel\_course}). Likewise, a student who is not enrolled
in any course can be deregistered (\C{deregister\_student}). Besides
\C{Student} and \C{Course} tables, there is also an \C{Enrollment}
table tracks the many-to-many enrollment relationship between courses
and students. The application is required to preserve the following
invariants on the database: 1.  $I_1$: An enrollment record should
always refer to an existing student and an existing course, and 2.
$I_2$: the capacity (\C{c\_capacity}) of a course should always be a
non-negative quantity. Both $I_1$ and $I_2$ can be violated under weak
isolation. $I_1$ can be violated, for example, when
\C{deregister\_student} runs concurrently with \C{enroll}, both at RC
isolation. While the former transaction removes the student record
after checking that no enrollments for that student exist, the latter
transaction concurrently adds an enrollment record after checking the
student exist. Both can succeed concurrently, resulting in an invalid
state. Invariant $I_2$ can be violated by two \C{enroll}s, both
reading the \C{c\_capacity}=1, and both (atomically) decrementing it,
resulting in \C{c\_capacity}=-1.  We ran \tool on the courseware
application written in our DSL, annotated with guarantees for all
transactions, and provided the $I = I_1 \wedge I_2$ as the correctness
condition. \tool was able to automatically discover anomalous
executions described above, and was subsequently able to infer that
the anomalies can be preempted by promoting the isolation level of
\C{enroll} and \C{deregister\_student} to SER (on both MySQL and
PostgreSQL), leaving the isolation levels of remaining transactions at
RC.

\begin{table}[]
\centering
\begin{tabular}{l|c|c|c|c|c|}
\cline{2-6}
                                 & \multicolumn{1}{l|}{\C{new\_order}} & \multicolumn{1}{l|}{\C{delivery}} & \multicolumn{1}{l|}{\C{payment}} & \multicolumn{1}{l|}{\C{order\_status}} & \multicolumn{1}{l|}{\C{stock\_level}} \\ \hline
\multicolumn{1}{|l|}{MySQL}      & SER                                   & SER                                 & RC                                 & RC                                       & RC                                      \\ \hline
\multicolumn{1}{|l|}{PostgreSQL} & RR                                    & RR                                  & RC                                 & RC                                       & RC                                      \\ \hline
\end{tabular}
\caption{The discovered isolation levels for TPC-C transactions}
\label{tab:tpcc}
\end{table}

\textbf{TPC-C} The simplified schema of the TPC-C benchmark has been
described in Sec.~\ref{sec:motivation}. In addition to the tables
shown in Fig.~\ref{fig:schema}, the TPC-C schema also has
\C{Warehouse} and \C{New\_order} tables that are relevant for
verification. To verify TPC-C, we consider four of the twelve
consistency conditions specified by the standard, which we name $I_1$
to $I_4$. Consistency condition $I_1$  requires that sales bottom line
of each warehouse equals to the sum of the sales bottom lines of all
district served by the warehouse.  Conditions $I_2$ and $I_3$
effectively enforce uniqueness of ids assigned to \C{Order} and
\C{New\_order} records, respectively, under a district. Condition
$I_4$ requires that the number of order lines under a district must
match the sum of order line counts of all orders under the district.
% There are five transactions in TPC-C, out of which two are read-only.
% The three remaining transactions are \C{new\_order}, \C{delivery}, and
% \C{payment}.  
Similar to the example discussed in Sec.~\ref{sec:motivation}, there
are a number of ways TPC-C's transactions violate the aforementioned
invariants under weak isolation. \tool was able to discover all such
violations when verifying the benchmark against $I =
\bigwedge_{i}I_i$, with guarantees of all three transactions
provided. The isolation levels were subsequently strengthened  as
shown in Table.~\ref{tab:tpcc}.

\begin{table*}[t]\small
\centering
\begin{tabular}{|c|c|c|c|}
  \hline
\textbf{Transaction}   & \textbf{Invariant} 
& \textbf{MySQL-Isolation} & \textbf{PostgreSQL-Isolation} \\ 
\hline
\multirow{4}{*}{\C{New\_Order} }  & $I_1$ 
& RC &  RC\\ 
&  $I_2$ &SER & RR \\
&  $I_3$ & SER  &  RR  \\
& $I_4$ & RC & RC   \\
\hline
\multirow{4}{*}{\C{Payment}}  & $I_1 $ 
& RC &  RC\\ 
&  $I_2$  &RC & RC \\
&  $I_3 $ & RC  &  RC  \\
& $I_4$  & RC & RC   \\
\hline
\multirow{4}{*}{\C{Delivery}}  & $I_1$  
& RC &  RC \\ 
&  $I_2$ &SER & RR \\
&  $I_3$ & SER  &  RR \\
& $I_4$  & RC & RC   \\
\hline
\end{tabular}
\caption{Various invariant violations witnessed for the TPC-C
  benchmark on MySQL and PostgreSQL}
\label{tab:tpcc-eval}
\end{table*}

To sanity-check the results of \tool, we conducted experiments with
the high-contention OLTP workload  on TPC-C aiming to explore the
space of correct isolation levels for different transactions. The
workload involves  a mix of all five TPC-C transactions executing
against a TPC-C database with 10 warehouses. Each warehouse has 10
districts, and each district serves 3000 customers. There are a total
of 5 transactions in TPC-C, and given that MySQL and PostgreSQL
support 3 isolation levels each, there are a total of $3^5 = 243$
different configurations of isolation levels for TPC-C transactions on
MySQL and PostgreSQL. We executed OLTP benchmark with all 243
configurations, and found 171 of them violating at least one invariant.
Table~\ref{fig:tpcc-eval} shows the summary of these results. As
expected, the isolation levels that \tool infers for the TPC-C
transactions do not result in invariant violations, either on MySQL or
on PostgreSQL.

### 2017-07-08 implementation.tex
Edits.

We have implemented our DSL to define transactions as monadic
computations in OCaml (modulo the syntactic sugar), and our automatic
reasoning framework as an extra frontend pass (called \tool) in ocamlc
4.03 compiler\footnote{The source code is available at available at
\emph{\ldots}}. The input to \tool is a program in our DSL that
describes the schema of the database as a collection of OCaml type
definitions, and transactions as OCaml functions, whose top-level
expression is an application of the \C{atomically\_do} combinator to a
function of type $a \rightarrow \C{DB}\;b$, for some (base) types $a$
and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema1} can
be described via the OCaml type definitions shown in
Fig.~\ref{fig:ocaml-schema}.  \tool also requires a specification of
the program in the form of a collection of guarantees ($G$), one per
transaction, and an invariant $I$ that is a conjunction of the
integrity constraints on the database. An auxiliary DSL that includes
the first-order logic (FOL) combinators has been implemented for this
purpose. \tool's verification pass follows OCaml's type checking pass,
hence the concrete artifact of verification is OCaml's typed AST. The
tool is already equipped with  an axiomatization of PostgresSQL and
MySQL's isolation levels expressed in our FOL DSL. Other data stores
can be similarly axiomatized. The concrete result of verification is
an assignment of an isolation level of the selected data store to each
transaction in the program.

At the top-level, \tool runs a loop that pics an unverified
transaction and progressively strengthens its isolation level until it
passes the verification. If the selected data store provides
serializable isolation level, and if the program is sequentially
corred, then the verification is guaranteed to succeed. Within the
loop, the \tool first computes the various rely relations needed for
verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
transaction, applying the inference rules to construct a state
tansformer, check its stability, and weaken it ($\stabilize{\cdot}$)
if it is not stable. The result of traversing the transaction's AST is
therefore a state transformer ($\F$) that is stable w.r.t $R_l$, which
is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
then checked against the transaction's stated guarantee ($G$). If the
check passes, then the guarantee is verified to check if it preserves
the invariant $I$. The successful result from both the checks results
in the transaction being certified correct under the current choice of
its isolation level. Succesful verification of all transactions
concludes the top-level execution, returning the inferred isolation
levels as its output.

@@

\textbf{Real-World Isolation Levels} The axiomatization of the
isolation levels presented in Sec.~\ref{sec:isolation} leave out
certain nuances of their implementations on real data stores, which
need to be taken into account for the verification to be effective in
practice.  We consider these into account while linking \tool with
store-specific semantics (isolation specifications etc). As an
example, consider how PostgresSQL implements the \C{UPDATE} operation.
\C{UPDATE} first selects the records that meet the search criteria
from the snapshot against which it is executing (the snapshot is
established at the beginning of the transaction if the isolation level
is RR, or at the beginning of the \C{UPDATE} statement if the
isolation level is RC). The selected records are then visited in the
actual database (if it still exists), write locks are obtained, and
the updatation is carried out, provided that the record still meets
\C{UPDATE}'s search criteria. If the record in no longer meets the
search criteria (due to a concurrent update) the record is excluded
from the update, and the write lock is immediately released.
Otherwise, the record remains locked till the transaction commits.

Clearly, the \C{UPDATE} operation on PostgreSQL is not atomic, as the
our formal model assumes it to be. It admits interference between the
updatation of individual records that meet the search criteria.
Nonetheless, through a series of relatively straightforward
deductions, we can show that PostgreSQL's \C{UPDATE} is infact
equivalent (in semantics) to a sequential composition of two atomic
operations $c_1;c_2$, where $c_1$ is effectively a \C{SELECT}
operation with the same search criteria as \C{UPDATE}, and $c_2$ is
the a slight variation of the original \C{UPDATE} that updates a
record only if a record with the same id is present in the records
returned by the \C{SELECT}. This transformation is summarized below:
\begin{smathpar}
\begin{array}{lcl}
\updatee{(\lambda x. ~e_1)}{(\lambda x.~e_2)}
&
\longrightarrow
&
\lete{y}{\selecte{(\lambda x.~e_1})}
     {\updatee{(\lambda x.~e_1 \wedge x.\idf\in\dom(y))}
              {(\lambda x.~e_2})}\\
\end{array}
\end{smathpar}
The intuition behind this translation is the observation that all
interferences possible during the execution of the \C{UPDATE} can be
accommodated between the time the records are selected from the
snapshot, and the time they are actually updated.  \tool performs this
translation if the selected store is PostgreSQL, allowing it to reason
about \C{UPDATE} operations in a way that is faithful to its semantics
on PostgreSQL. \tool also admits similar compensatory logic for
certain combinations of isolation levels and operations on MySQL.

@@

\textbf{Annotation Burden} \tool significantly reduces the annotation
burden in verifying a weakly isolated transactions by eliminating the
need to annotate intermediary assertions and loop invariants.
Guarantees ($G$) and invariants ($I$), however, still need to be
provided. Alternatively, a weakly isolated transaction $T$ can be
verified against the generic serializability condition, to eliminate
the need for any annotation. In this mode, \tool first infers the
transformer $\F_{SER}$ of $T$ without considering any interference,
which then becomes its guarantee ($G$). Doing likewise for every
transaction results in a rely relation ($R$) that includes $\F_{SER}$
of every transaction. Verification now proceeds by taking interference
into account, and verifying that each transaction still yeilds the
same $F$ as its $F_{SER}$. The result of this verification is an
assignment of (possibly weak) isolation levels to transactions which
nonetheless guarantees serializability of all executions.

>>>>

We have implemented our DSL to define transactions as monadic
computations in OCaml (modulo the syntactic sugar), and our automatic
reasoning framework as an extra frontend pass (called \tool) in ocamlc
4.03 compiler\footnote{The source code is available at available at
\emph{\ldots}}. The input to \tool is a program in our DSL that
describes the schema of the database as a collection of OCaml type
definitions, and transactions as OCaml functions, whose top-level
expression is an application of the \C{atomically\_do} combinator to a
function of type $a \rightarrow \C{DB}\;b$, for some (base) types $a$
and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema1} can
be described via the OCaml type definitions shown in
Fig.~\ref{fig:ocaml-schema}.  \tool also requires a specification of
the program in the form of a collection of guarantees ($G$), one per
transaction, and an invariant $I$ that is a conjunction of the
integrity constraints on the database. An auxiliary DSL that includes
the first-order logic (FOL) combinators has been implemented for this
purpose. \tool's verification pass follows OCaml's type checking pass,
hence the concrete artifact of verification is OCaml's typed AST. The
tool is already equipped with  an axiomatization of PostgreSQL and
MySQL's isolation levels expressed in our FOL DSL. Other data stores
can be similarly axiomatized. The concrete result of verification is
an assignment of an isolation level of the selected data store to each
transaction in the program.

At the top-level, \tool runs a loop that pics an unverified
transaction and progressively strengthens its isolation level until it
passes the verification. If the selected data store provides
serializable isolation level, and if the program is sequentially
correct, then the verification is guaranteed to succeed. Within the
loop, the \tool first computes the various rely relations needed for
verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
transaction, applying the inference rules to construct a state
transformer, check its stability, and weaken it ($\stabilize{\cdot}$)
if it is not stable. The result of traversing the transaction's AST is
therefore a state transformer ($\F$) that is stable w.r.t $R_l$, which
is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
then checked against the transaction's stated guarantee ($G$). If the
check passes, then the guarantee is verified to check if it preserves
the invariant $I$. The successful result from both the checks results
in the transaction being certified correct under the current choice of
its isolation level. Successful verification of all transactions
concludes the top-level execution, returning the inferred isolation
levels as its output.

@@

\textbf{Real-World Isolation Levels} The axiomatization of the
isolation levels presented in Sec.~\ref{sec:isolation} leave out
certain nuances of their implementations on real data stores, which
need to be taken into account for the verification to be effective in
practice.  We consider these into account while linking \tool with
store-specific semantics (isolation specifications etc). As an
example, consider how PostgreSQL implements the \C{UPDATE} operation.
\C{UPDATE} first selects the records that meet the search criteria
from the snapshot against which it is executing (the snapshot is
established at the beginning of the transaction if the isolation level
is RR, or at the beginning of the \C{UPDATE} statement if the
isolation level is RC). The selected records are then visited in the
actual database (if it still exists), write locks are obtained, and
the updatation is carried out, provided that the record still meets
\C{UPDATE}'s search criteria. If the record in no longer meets the
search criteria (due to a concurrent update) the record is excluded
from the update, and the write lock is immediately released.
Otherwise, the record remains locked till the transaction commits.

Clearly, the \C{UPDATE} operation on PostgreSQL is not atomic, as the
our formal model assumes it to be. It admits interference between the
updatation of individual records that meet the search criteria.
Nonetheless, through a series of relatively straightforward
deductions, we can show that PostgreSQL's \C{UPDATE} is in fact
equivalent (in semantics) to a sequential composition of two atomic
operations $c_1;c_2$, where $c_1$ is effectively a \C{SELECT}
operation with the same search criteria as \C{UPDATE}, and $c_2$ is
the a slight variation of the original \C{UPDATE} that updates a
record only if a record with the same id is present in the records
returned by the \C{SELECT}. This transformation is summarized below:
\begin{smathpar}
\begin{array}{lcl}
\updatee{(\lambda x. ~e_1)}{(\lambda x.~e_2)}
&
\longrightarrow
&
\lete{y}{\selecte{(\lambda x.~e_1})}
     {\updatee{(\lambda x.~e_1 \wedge x.\idf\in\dom(y))}
              {(\lambda x.~e_2})}\\
\end{array}
\end{smathpar}
The intuition behind this translation is the observation that all
interferences possible during the execution of the \C{UPDATE} can be
accommodated between the time the records are selected from the
snapshot, and the time they are actually updated.  \tool performs this
translation if the selected store is PostgreSQL, allowing it to reason
about \C{UPDATE} operations in a way that is faithful to its semantics
on PostgreSQL. \tool also admits similar compensatory logic for
certain combinations of isolation levels and operations on MySQL.

@@

\textbf{Annotation Burden} \tool significantly reduces the annotation
burden in verifying a weakly isolated transactions by eliminating the
need to annotate intermediary assertions and loop invariants.
Guarantees ($G$) and invariants ($I$), however, still need to be
provided. Alternatively, a weakly isolated transaction $T$ can be
verified against the generic serializability condition, to eliminate
the need for any annotation. In this mode, \tool first infers the
transformer $\F_{SER}$ of $T$ without considering any interference,
which then becomes its guarantee ($G$). Doing likewise for every
transaction results in a rely relation ($R$) that includes $\F_{SER}$
of every transaction. Verification now proceeds by taking interference
into account, and verifying that each transaction still yields the
same $F$ as its $F_{SER}$. The result of this verification is an
assignment of (possibly weak) isolation levels to transactions which
nonetheless guarantees serializability of all executions.
